---
title: "Assignment 10: Data Scraping"
author: "Trisha Belus"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)

getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
website <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Water_system <-website %>% html_nodes ("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
PWSID <-website %>% html_nodes ("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
Ownership <-website %>% html_nodes ("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
MGD_months <- website %>% html_nodes ('th~ td+ td , th~ td+ td') %>% html_text()
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

df_deq <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12), 
                             "Year" = rep(2024,12),
                             "Maximum_Day_Use" = as.numeric(MGD_months)) %>% 
  mutate(Water_System_Name = !!Water_system,
         PWSID = !!PWSID,
         Ownership = !!Ownership,
         Date = my(paste(Month,"-",Year)))

#5 
ggplot(df_deq,aes(x=Date,y=Maximum_Day_Use)) + 
  geom_line() 

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 

scrape.it <- function(the_year, the_pwsid){
  
  base_url <- "https://www.ncwater.org/WUDC/app/LWSP/report.php"
  scrape_website <- read_html(paste0(base_url, '?pwsid=', the_pwsid, '&year=', the_year))
  
  Water_system_tag <-"div+ table tr:nth-child(1) td:nth-child(2)"
  PWSID_tag <-"td tr:nth-child(1) td:nth-child(5)"
  Ownership_tag <-"div+ table tr:nth-child(2) td:nth-child(4)"
  MGD_months_tag <- "th~ td+ td , th~ td+ td"
  
  the_water_system <- scrape_website %>% html_nodes(Water_system_tag) %>% html_text()
  the_pwsid_name <- scrape_website %>%   html_nodes(PWSID_tag) %>%  html_text()
  the_ownership_name <- scrape_website %>% html_nodes(Ownership_tag) %>% html_text()
  the_mgd_value <- scrape_website %>% html_nodes(MGD_months_tag) %>% html_text()
  
  df_scraped <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12), 
                             "Year" = rep(the_year,12),
                             "Maximum_Day_Use" = as.numeric(the_mgd_value)) %>% 
  mutate(Water_System_Name = !!the_water_system,
         PWSID = !!the_pwsid_name,
         Ownership = !!the_ownership_name,
         Date = my(paste(Month,"-",Year)))
  
  Sys.sleep(1) 

  return(df_scraped)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
durham_2020 <- scrape.it(2020,"03-32-010")
glimpse(durham_2020)

ggplot(durham_2020)+
  geom_line(aes(x = Date, y = Maximum_Day_Use))

```

8. Use the function above to extract data for Cary (PWSID = '03-92-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
cary_2020 <- scrape.it(2020,"03-92-020")
glimpse(cary_2020)


combined_df <- rbind(durham_2020,cary_2020)

ggplot(data = combined_df)+
  geom_line(aes(x=Date, y= Maximum_Day_Use, color = Water_System_Name))
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
cary_18_23 <- scrape.it(2018:2023,"03-92-020")
glimpse(cary_18_23)

the_years = rep(2018:2023)
cary_pwsid = "03-92-020"

the_dfs <- map(the_years,scrape.it,the_pwsid=cary_pwsid)
the_df <- bind_rows(the_dfs)

ggplot(the_df,aes(x=Date,y=Maximum_Day_Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE)

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Just by looking at the plot, Cary has a positive upwards trend in water usage over time as shown by the smoothed line on the graph.
>
